`methyvim`: an R package for Variable Importance Measures for

### 2017-10-18 paper/paper_BiocF1000.Rmd
minor formatting
f1000 paper edits, minor cleanup

---
title: '`methyvim`: an R package for Variable Importance Measures for
Differential Methylation Analysis'
author:
  - name: Nima S. Hejazi
    affiliation: Address of author1
  - name: Rachael V. Phillips
    affiliation: orchid.org/0000-0002-8474-591X
  - name: Alan E. Hubbard
    affiliation: Address of author2
  - name: Mark J. van der Laan
    affiliation: Address of author2
abstract: |
  We present `methyvim`, a general algorithm for the nonparametric estimation of
  treatment/exposure effects on DNA methylation at CpG sites, complete with straightforward statistical inference for such estimates. This approach
  leverages variable importance measures, a class of parameters that arise in
  causal inference and are defined in such a manner that they provide targeted
  estimates of the relative importance of individual CpG site with respect to
  a binary exposure/treatment assigned at the level of subjects. Additionally,
  this procedure is computationally efficient; incorporating a preliminary
  screening step to isolate a subset of sites for which there is cursory
  evidence of differential methylation and a multiple testing correction step
  to control the False Discovery Rate as if all sites were tested. This
  technique for analysis of differentially methylated positions provides an
  avenue to move beyond linear models without the loss of interpretable
  statistical inference.
keywords: DNA methylation, differential methylation, epigenetics, causal
  inference, variable importance, targeted minimum loss-based estimation
bibliography: paper_BiocF1000.bib
output: BiocWorkflowTools::f1000_article
---

@@

DNA methylation is a fundamental epigenetic process known to play an important
role in the regulation of gene expression. DNA CpG methylation involves the
addition of a methyl group ($\text{CH}_3$) to the fifth carbon of the cytosine
ring structure to form 5-methylcytosine. Numerous biological and medical studies
have implicated DNA CpG methylation as playing a role in disease and
development [@robertson2005dna]. Perhaps unsurprisingly then, biotechnologies
have been developed to study the molecular mechanisms of this epigenetic
process. Modern assays, like the Illumina _Infinium_ Methylation assay,
allow for quantitative interrogation of DNA methylation of CpG sites scattered
across the genome at single-nucleotide resolution; moreover, much effort
has been invested, by the bioinformatics community, in the development of tools
for properly removing technological effects that may contaminate biological
signatures measured by such assays. Despite these advances in both biological
and bioninformatical techniques, most statistical methods available for
differential analysis of data produced by such assays rely on over-simplified
(often generalized linear) models. This is because typical statistical
estimation approaches are not achievable for such high-dimensional data without restrictive modeling assumptions. When these assumptions are violated, estimates
of the population-level effect of an exposure or treatment may have large bias.
Data-adaptive estimation procedures that utilize machine learning can control
for high-dimensional confounders when estimating treatment/exposure effects.
However, interpretable statistical inference (i.e. confidence intervals,
hypothesis tests) resulting from these data-adaptive estimates is challenging [@libbrecht2015machine]. In this paper, we present an alternative such
statistical analysis approaches in the form of nonparametric estimation
procedures that provide simple and well interpretable statistical inference.
Inspired by concepts of counterfactual effects from the causal inference
literature and machine learning algorithms, we provide a computationally
efficient technique for obtaining targeted estimates of nonparametric _variable
importance measures_ (__VIM__) for a pre-screened set of CpGs, controlling for
the False Discovery Rate (FDR) as if all sites were tested. Under counterfactual assumptions, these estimators have asymptotic normal sampling distributions
and thus provide reliable inference. Counterfactuals are "counter"-fact, defined
in terms of an unrealistic yet ideal experiment. In the context of DNA
methylation and for a binary treatment, we define the counterfactual
outcomes as the Beta-value/M-value a CpG site would have if all subjects were
treated and the Beta-value/M-value a CpG site would have if all subjects were
not treated. Of course, these counterfactual outcomes are impossible to observe
but they are very helpful for us to asses the importance of a CpG site relative
to a subjects treatment/exposure and they can be statistically estimated [@van2006statistical]. Our VIM definition uses the concepts of counterfactuals
and the framework for estimation is built on _targeted maximum likelihood
estimation or targeted minimum loss-based estimation_ (__TMLE__)
[@van2006targeted]. This methodology targets the individual importance of each
CpG site by taking advantage of the prediction power of machine learning
algorithms. The mathematical properties of this approach have been explored in
[@tuglus2008targeted] and we focus on our contribution to it's application,
`methyvim`, an R package that accommodates this technique for differential
methylation analysis.

@@

1. Pre-screening of genomic sites is used to isolate a subset of sites for
  which there is cursory evidence of differential methylation. For the sake of
  computational feasibility, targeted minimum loss-based estimates of VIMs are
  computed only for this subset of sites. Currently, the available screening
  approach adapts core routines from the
  [`limma`](http://bioconductor.org/packages/release/bioc/html/limma.html) R
  package. Future releases will support functionality from other packages (e.g., [`randomForest`](https://cran.r-project.org/web/packages/randomForest/index.html),
  [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi)). Following the
  style of the function for performing screening via `limma`, users may write
  their own screening functions and are invited to contribute such functions to
  the core software package by opening pull requests at the GitHub repository.

2. Nonparametric estimates of VIMs, for the specified target parameter, are
  computed at each of the CpG sites passing the screening step. The VIMs are
  defined in such a way that the estimated effects is of an exposure/treatment
  on the methylation status of a target CpG site, controlling for the observed
  methylation status of the neighbors of that site. Currently, routines are
  adapted from the [`tmle`](https://CRAN.R-project.org/package=tmle) R package.
  Future releases will support doubly-robust estimates of these VIMs (via the
  [`drtmle`](https://cran.r-project.org/web/packages/drtmle/index.html) package)
  and add parameters for continuous treatments/exposures (via the
  [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi) package).

@@

#### Class methytmle

@@

Software available from:

Latest source code: https://github.com/nhejazi/methyvim

@@

Software license:

@@

NH designed and implemented the tool, applied the tool to the use case, and
drafted the vignette. RP helped design the tool and drafted the article.
AH and ML serve as the advisors for the development of this tool.

@@

The author(s) declared that no grants were involved in supporting this work.

>>>>

---
title: '`methyvim`: an R package for Variable Importance Measures for
Differential Methylation Analysis'
author:
  - name: Nima S. Hejazi
    affiliation: 1
    orcid: 0000-0002-7127-2789
  - name: Rachael V. Phillips
    affiliation: 1
    orcid: 0000-0002-8474-591X
  - name: Alan E. Hubbard
    affiliation: 1
    orcid: 0000-0002-3769-0127
  - name: Mark J. van der Laan
    affiliation: 1, 2
    orcid: TBA
affiliations:
  - name: Division of Biostatistics, University of California, Berkeley
    index: 1
  - name: Department of Statistics, University of California, Berkeley
    index: 2
abstract: |
  We present `methyvim`, a general algorithm for the nonparametric estimation of
  treatment/exposure effects on DNA methylation at CpG sites, complete with
  straightforward statistical inference for such estimates. This approach
  leverages variable importance measures, a class of parameters that arise in
  causal inference and are defined in such a manner that they provide targeted
  estimates of the relative importance of individual CpG site with respect to
  a binary exposure/treatment assigned at the level of subjects. Additionally,
  this procedure is computationally efficient; incorporating a preliminary
  screening step to isolate a subset of sites for which there is cursory
  evidence of differential methylation and a multiple testing correction step
  to control the False Discovery Rate as if all sites were tested. This
  technique for analysis of differentially methylated positions provides an
  avenue to move beyond linear models without the loss of interpretable
  statistical inference.
keywords: DNA methylation, differential methylation, epigenetics, causal
  inference, variable importance, targeted minimum loss-based estimation
bibliography: paper_BiocF1000.bib
output: BiocWorkflowTools::f1000_article
---

@@

DNA methylation is a fundamental epigenetic process known to play an important
role in the regulation of gene expression. DNA CpG methylation involves the
addition of a methyl group ($\text{CH}_3$) to the fifth carbon of the cytosine
ring structure to form 5-methylcytosine. Numerous biological and medical studies
have implicated DNA CpG methylation as playing a role in disease and
development [@robertson2005dna]. Perhaps unsurprisingly then, biotechnologies
have been developed to study the molecular mechanisms of this epigenetic
process. Modern assays, like the Illumina _Infinium_ Methylation assay,
allow for quantitative interrogation of DNA methylation of CpG sites scattered
across the genome at single-nucleotide resolution; moreover, much effort
has been invested, by the bioinformatics community, in the development of tools
for properly removing technological effects that may contaminate biological
signatures measured by such assays. Despite these advances in both biological
and bioninformatical techniques, most statistical methods available for
differential analysis of data produced by such assays rely on over-simplified
(often generalized linear) models. This is because typical statistical
estimation approaches are not achievable for such high-dimensional data without
restrictive modeling assumptions. When these assumptions are violated, estimates
of the population-level effect of an exposure or treatment may have large bias.
Data-adaptive estimation procedures that utilize machine learning can control
for high-dimensional confounders when estimating treatment/exposure effects.
However, interpretable statistical inference (i.e. confidence intervals,
hypothesis tests) resulting from these data-adaptive estimates is challenging
[@libbrecht2015machine]. In this paper, we present an alternative such
statistical analysis approaches in the form of nonparametric estimation
procedures that provide simple and well interpretable statistical inference.
Inspired by concepts of counterfactual effects from the causal inference
literature and machine learning algorithms, we provide a computationally
efficient technique for obtaining targeted estimates of nonparametric _variable
importance measures_ (__VIM__) for a pre-screened set of CpGs, controlling for
the False Discovery Rate (FDR) as if all sites were tested. Under
counterfactual assumptions, these estimators have asymptotic normal sampling
distributions and thus provide reliable inference. Counterfactuals are
"counter"-fact, defined in terms of an unrealistic yet ideal experiment. In the
context of DNA methylation and for a binary treatment, we define the
counterfactual outcomes as the Beta-value/M-value a CpG site would have if all
subjects were treated and the Beta-value/M-value a CpG site would have if all
subjects were not treated. Of course, these counterfactual outcomes are
impossible to observe but they are very helpful for us to asses the importance
of a CpG site relative to a subjects treatment/exposure and they can be
statistically estimated [@vdl2006statistical]. Our VIM definition uses the
concepts of counterfactuals and the framework for estimation is built on
_targeted maximum likelihood estimation_ or _targeted minimum loss-based
estimation_ (__TMLE__) [@vdl2006targeted]. This methodology targets the
individual importance of each CpG site by taking advantage of the prediction
power of machine learning algorithms. The mathematical properties of this
approach have been explored in [@tuglus2008targeted] and we focus on our
contribution to it's application, `methyvim`, an R package that make this
technique for differential methylation analysis easily accessible.

@@

1. Pre-screening of genomic sites is used to isolate a subset of sites for
   which there is cursory evidence of differential methylation. For the sake of
   computational feasibility, targeted minimum loss-based estimates of VIMs are
   computed only for this subset of sites. Currently, the available screening
   approach adapts core routines from the
   [`limma`](http://bioconductor.org/packages/limma) R
   package. Future releases will support functionality from other packages
   (e.g., [`randomForest`](https://CRAN.R-project.org/package=randomForest),
   [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi)). Following the
   style of the function for performing screening via `limma`, users may write
   their own screening functions and are invited to contribute such functions to
   the core software package by opening pull requests at the GitHub repository.

2. Nonparametric estimates of VIMs, for the specified target parameter, are
   computed at each of the CpG sites passing the screening step. The VIMs are
   defined in such a way that the estimated effects is of an exposure/treatment
   on the methylation status of a target CpG site, controlling for the observed
   methylation status of the neighbors of that site. Currently, routines are
   adapted from the [`tmle`](https://CRAN.R-project.org/package=tmle) R package.
   Future releases will support doubly-robust estimates of these VIMs (via the
   [`drtmle`](https://cran.r-project.org/package=drtmle) package)
   and add parameters for continuous treatments/exposures (via the
   [`tmle.npvi`](https://CRAN.R-project.org/package=tmle.npvi) package).

@@

#### Class `methytmle`

@@

Latest source code (development version): https://github.com/nhejazi/methyvim

Bioconductor (stable release): https://bioconductor.org/packages/methyvim

@@

https://github.com/nhejazi/methyvim

Software license: MIT, copyright Nima S. Hejazi

@@

NH designed and implemented the software package, applied the tool to the use
cases, and co-drafted the present article. RP helped in designing the software
and co-drafted the present article. AH and ML serve as the advisors for the
development of this software and general statistical algorithm.

@@

NH was supported in part by the National Library Of Medicine of the National
Institutes of Health under Award Number T32LM012417, by [SUPERFUND GRANT], and
by [NINA HOLLAND GRANT]. RP was supported by [SUPERFUND GRANT]. The content of
this work is solely the responsibility of the authors and does not necessarily
represent the official views of the various funding sources and agencies.
